Decoding Mortality: Trends and Causes in Canada between 2001-2022

### 2024-03-17 outputs/paper/paper.qmd
Updated the llm folder

---
title: "Decoding Mortality: Trends and Causes in Canada between 2001-2022"
subtitle: "A Bayesian Statistical Analysis Unveiling the Predominance of Malignant Neoplasms and Heart Diseases"
author: 
  - Yuchao Niu
thanks: "Code and data are available at: https://github.com/MelanieNiu/Canadian-Mortality."
date: yesterday
date-format: long
abstract: "The study of mortality offers key public health insights into the fundamental causes and trends of death. This research investigates Canada's mortality trends from 2001 to 2022, employing Bayesian Poisson and negative binomial regression models to analyze data from the Canadian Vital Statistics Death Database. It reveals that malignant neoplasms and heart diseases significantly elevate mortality rates, surpassing the effects of unintentional accidents by five times and three times, respectively. Such findings highlight the critical necessity for targeted public health initiatives focusing on the prevention, early detection, and treatment of these conditions."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

library(tidyverse)
library(knitr)
library(rstanarm)
library(modelsummary)
library(gt)
library(arrow)
library(dplyr)
library(broom.mixed)

@@

This study investigates mortality in Canada from 2001 to 2022, utilizing the Canadian Vital Statistics Death Database by Statistics Canada. The focus is on identifying the risk factors associated with each cause of death. We employed a Bayesian approach, using both the Poisson regression model and the Negative Binomial regression model, to analyze the top five leading causes of mortality in Canada. The Bayesian model applies Bayesian inference to continually update the probability of a hypothesis as more data becomes available. The Poisson regression model is traditionally favored for count data but assumes equal mean and variance, an assumption that does not always hold in mortality data due to over-dispersion. The Negative Binomial regression model addresses this by accounting for over-dispersion, offering a more flexible approach for such data.

@@

```{r}
#| label: fig-2
#| fig-cap: Number of deaths from the top-five causes % of all deaths in 2021
#| echo: false
#| warning: false
#| message: false

# cleaned_data |>
#   filter(Year == 2021) |>
#   mutate(Percentage = Death / sum(Death) * 100) |>
#   ggplot(top_five, aes(x="", y= Death, fill= Cause)) +
#   geom_bar(width = 1, stat = "identity") +
#   coord_polar("y") +
#   theme_void() +
#   labs(fill="Cause of Death")

@@

```{r}
#|message= false
#|warning= false
#|include= false

@@

coef_short_names_corrected <- c(
  "(Intercept)" = "(Intercept)",
  "CauseCerebrovascular diseases" = "Cerebrovascular diseases",
  "CauseChronic lower respiratory diseases" = "Chronic lower respiratory diseases",
  "CauseDiseases of heart" = "Diseases of heart",
  "CauseMalignant neoplasms" = "Malignant neoplasms",
  "CauseCOVID-19" = "COVID-19"  # Ensure this matches the term exactly as it appears in your model output
)

# Apply the corrected mapping to rename terms for the Poisson model
tidy_poisson$term <- coef_short_names_corrected[tidy_poisson$term]
# Handling any terms that did not match by retaining their original names
tidy_poisson$term[is.na(tidy_poisson$term)] <- names(coef_short_names_corrected)[is.na(tidy_poisson$term)]

# Apply the corrected mapping to rename terms for the Negative Binomial model
tidy_negbinomial$term <- coef_short_names_corrected[tidy_negbinomial$term]
# Handling any terms that did not match by retaining their original names
tidy_negbinomial$term[is.na(tidy_negbinomial$term)] <- names(coef_short_names_corrected)[is.na(tidy_negbinomial$term)]

# Combine the data for comparison
comparison <- bind_rows(
  mutate(tidy_poisson, model = "Poisson"),
  mutate(tidy_negbinomial, model = "Negative Binomial")
)

# Assuming 'comparison' is your final data frame from previous steps
comparison_table <- comparison %>%
  gt() %>%
  tab_header(
    title = "Model Comparison: Poisson vs. Negative Binomial"
  ) %>%
  cols_label(
    term = "Term",
    estimate = "Estimate",
    std.error = "Standard Error",
    model = "Model"
  ) %>%
  fmt_number(
    columns = c(estimate, std.error),
    decimals = 3
  ) %>%
  tab_spanner(
    label = "Statistics",
    columns = c(estimate, std.error)
  ) %>%
  tab_style(
    style = cell_text(align = 'center'),
    locations = cells_body(columns = everything())
  ) %>%
  tab_options(
    heading.title.font.size = 14,
    heading.subtitle.font.size = 12,
    table.font.size = 12
  )
```

@@

comparison_table
```

For instance, @tbl-3 shows that diseases of the heart have a coefficient of 1.454 in both models, suggesting the expected death count from this cause is approximately `r round(exp(1.454),2)`  times higher than from Accidents (unintentional injuries), after exponentiating the coefficient. Similarly, malignant neoplasms have a coefficient of approximately 1.805, indicating an expected count more than `r round(exp(1.805)-1, 2)` times higher than Accidents. Conversely, chronic lower respiratory diseases show a modest decrease in mortality compared to Accidents, with a coefficient of -0.070 in the Poisson model, indicating about `r round(1-exp(-0.070),2)` fewer deaths.

Table shows that among the top five leading causes of death in Canada, diseases of the heart with a coefficient of 1.454 and malignant neoplasms with a coefficient of approximately 1.805 significantly indicates a strong positive association in both models. The coefficient for Diseases of heart is 1.454 in both the Poisson model and the negative binomial model. Exponentiating the coefficient 1.454 to obtain suggests that the expected count of deaths due to Diseases of the heart is about `r round(exp(1.454) - 1,2)` times higher than the baseline category Accidents (unintentional injuries). Similarly, exponentiating the coefficient for malignant neoplasms to obtain `r round(exp(1.805),2)` shows the expected count of deaths is ` times higher Accidents (unintentional injuries). Conversely, chronic lower respiratory diseases are associated with a slight decrease in mortality compared to Accidents (unintentional injuries) with a coefficient of -0.070 in the Poisson model, although the impact is modest about  less.

Both models yield similar coefficients, suggesting they capture a similar relationship between causes and death counts. The negative binomial model, with a log likelihood of -1050.577, fits the data better than the Poisson model, which has a log likelihood of -16957.794. The negative binomial model's better fit is further supported by a higher ELPD value compared to the Poisson model as shown in @tbl-4. Posterior predictive checks, trace plots, and Rhat plots assess MCMC convergence and the reliability of the Bayesian analysis, with details in @sec-model-details.

```{r}
#| label: tbl-4
#| tbl-cap: "Cross validation comparision between Poisson model and negative binomial model "
#| echo: false
#| warning: false
#| message: false

@@

(loo_compare(poisson, neg_binomial))

@@

@@

Future studies could aim to address these weaknesses by incorporating spatial and demographic variables to account for clustering effects. Future studies can broaden the scope to include more regions, socioeconomical variables, and their interactions. Longitudinal studies could offer insights into mortality trends and the long-term effects of environmental factors like climate change.

@@

The raw dataset was downloaded from Statistics Canada open data portal and cleaned by selecting the variables of interest "Year", "Causes of Death", "Characteristics" and "Value". The "Characteristics" variable is converted into a new variable "Death" indicating the number of death attributed to the cause and a new "Rank" variable showing the rank of the cause in a given year. @tbl-5 presents a sample of the cleaned dataset.

```{r}
#| label: tbl-5
#| tbl-cap: "A sample of the cleaned dataset of number of death attributed to causes of death in Canada, 2001-2022"
#| echo: false
#| warning: false
#| message: false
sampled_data <- sample_n(cleaned_data, 5) %>%
  kable()
sampled_data
```

>>>>

---
title: "Decoding Mortality: Trends and Causes in Canada between 2001-2022"
subtitle: "A Bayesian Statistical Analysis Unveiling the Predominance of Malignant Neoplasms and Heart Diseases"
author: 
  - Yuchao Niu
thanks: "Code and data are available at: https://github.com/MelanieNiu/Canadian-Mortality."
date: today
date-format: long
abstract: "The study of mortality offers key public health insights into the fundamental causes and trends of death. This research investigates Canada's mortality trends from 2001 to 2022, employing Bayesian Poisson and negative binomial regression models to analyze data from the Canadian Vital Statistics Death Database. It reveals that malignant neoplasms and heart diseases significantly elevate mortality rates, surpassing the effects of unintentional accidents by five times and three times, respectively. Such findings highlight the critical necessity for targeted public health initiatives focusing on the prevention, early detection, and treatment of these conditions."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

library(tidyverse)
library(knitr)
library(rstanarm)
library(modelsummary)
library(gt)
library(arrow)
library(dplyr)
library(broom.mixed)
library(loo)

@@

This study investigates mortality in Canada from 2001 to 2022, utilizing the Canadian Vital Statistics Death Database by Statistics Canada. The estimand of the study is the relative risk factors for death associated with each cause of death. We employed a Bayesian approach, using both the Poisson regression model and the Negative Binomial regression model, to analyze the top five leading causes of mortality in Canada. The Bayesian model applies Bayesian inference to continually update the probability of a hypothesis as more data becomes available. The Poisson regression model is traditionally favored for count data but assumes equal mean and variance, an assumption that does not always hold in mortality data due to over-dispersion. The Negative Binomial regression model addresses this by accounting for over-dispersion, offering a more flexible approach for such data. Through this study we also would like to compare the performance of the two models for the dataset we are interested in.

@@

@@

```{r}
#| echo: false
#| warning: false
#| message: false

@@

coef_short_names_corrected <- c(
  "(Intercept)" = "(Intercept)",
  "CauseCerebrovascular diseases" = "Cerebrovascular diseases",
  "CauseChronic lower respiratory diseases" = "Chronic lower respiratory diseases",
  "CauseDiseases of heart" = "Diseases of heart",
  "CauseMalignant neoplasms" = "Malignant neoplasms",
  "CauseCOVID-19" = "COVID-19"  
)

tidy_poisson$term <- coef_short_names_corrected[tidy_poisson$term]

tidy_poisson$term[is.na(tidy_poisson$term)] <- names(coef_short_names_corrected)[is.na(tidy_poisson$term)]

tidy_negbinomial$term <- coef_short_names_corrected[tidy_negbinomial$term]
tidy_negbinomial$term[is.na(tidy_negbinomial$term)] <- names(coef_short_names_corrected)[is.na(tidy_negbinomial$term)]

comparison <- bind_rows(
  mutate(tidy_poisson, model = "Poisson"),
  mutate(tidy_negbinomial, model = "Negative Binomial")
)

comparison_table <- comparison %>%
  gt() %>%
  tab_header(
    title = "Model Comparison: Poisson vs. Negative Binomial"
  ) %>%
  cols_label(
    term = "Term",
    estimate = "Estimate",
    std.error = "Standard Error",
    model = "Model"
  ) %>%
  fmt_number(
    columns = c(estimate, std.error),
    decimals = 3
  ) %>%
  tab_spanner(
    label = "Statistics",
    columns = c(estimate, std.error)
  ) %>%
  tab_style(
    style = cell_text(align = 'center'),
    locations = cells_body(columns = everything())
  ) %>%
  tab_options(
    heading.title.font.size = 14,
    heading.subtitle.font.size = 12,
    table.font.size = 12
  )

# LOO for both models
loo_poisson <- loo(canada_death_poisson)
loo_negbinomial <- loo(canada_death_negbinomial)

# Assuming loo_poisson and loo_negbinomial have been calculated as shown:
loo_poisson <- loo(canada_death_poisson, cores = 2)
loo_negbinomial <- loo(canada_death_negbinomial, cores = 2)

# Accessing LOOIC and ELPD correctly
looic_poisson <- loo_poisson$looic
elpd_loo_poisson <- loo_poisson$pointwise[, "elpd_loo"]
elpd_se_poisson <- sqrt(var(elpd_loo_poisson))

looic_negbinomial <- loo_negbinomial$looic
elpd_loo_negbinomial <- loo_negbinomial$pointwise[, "elpd_loo"]
elpd_se_negbinomial <- sqrt(var(elpd_loo_negbinomial))

model_summary <- data.frame(
  Metric = c("LOOIC", "ELPD", "ELPD SE"),
  Poisson = c(looic_poisson, sum(elpd_loo_poisson), elpd_se_poisson),
  Negative_Binomial = c(looic_negbinomial, sum(elpd_loo_negbinomial), elpd_se_negbinomial)
)
```

@@

comparison_table

For instance, @tbl-3 shows that diseases of the heart have a coefficient of 1.454 in both models, suggesting the expected death count from this cause is approximately `r round(exp(1.454),2)`  times higher than from Accidents (unintentional injuries), after exponentiating the coefficient. Similarly, malignant neoplasms have a coefficient of approximately 1.805, indicating an expected count more than `r round(exp(1.805)-1, 2)` times higher than Accidents. Chronic lower respiratory diseases show a modest increase in mortality compared to Accidents, with a coefficient of 0.320 in the Poisson model and 0.322 in the negative binomial model, indicating about `r round(exp(0.320)-1,2)` times more deaths than Accidents. Lastly, Cerebrovascular diseases has a coefficient of 0.141, showing `r round(exp(0.141)-1,2)` times more death counts compare to Accidents although the impact is also modest.

@tbl-4 displayed the values of the parameters in both models.The LOOIC value (Leave-One-Out Cross-Validation Information Criterion) is useful in comparing model fitness. A lower LOOIC value suggests a model that is expected to make more accurate predictions for new, unseen data. This is because LOOIC penalizes models for having too many parameters (complexity), helping to avoid overfitting. The Poisson model has a LOOIC value of 33604.96 whereas the negative binomial model has a LOOIC value of 1787.36 much lower than the Poisson model. This indicates a better fit of the data by the negative binomial model.

```{r}
#| label: tbl-4
#| tbl-cap: "Model summary of Poisson model vs. negative binomial model"
#| echo: false
#| warning: false
#| message: false

model_summary %>%
  gt() %>%
  tab_header(
    title = "Model Summary: Poisson vs. Negative Binomial"
  ) %>%
  cols_label(
    Metric = "Metric",
    Poisson = "Poisson",
    Negative_Binomial = "Negative Binomial"
  ) %>%
  fmt_number(
    columns = vars(Poisson, Negative_Binomial),
    decimals = 2
  )
```

Both models yield similar coefficients, suggesting they capture a similar relationship between causes and death counts. The negative binomial model, with a log likelihood of -1050.577, fits the data better than the Poisson model, which has a log likelihood of -16957.794. The negative binomial model's better fit is further supported by a higher ELPD value compared to the Poisson model as shown in @tbl-5. Posterior predictive checks, trace plots, and Rhat plots assess MCMC convergence and the reliability of the Bayesian analysis, with details in @sec-model-details.
\newpage

```{r}
#| label: tbl-5
#| tbl-cap: "Cross validation comparision between Poisson model and negative binomial model "
#| echo: false
#| warning: false
#| message: false

@@

comparison_result <- loo_compare(poisson, neg_binomial)
comparison_df <- as.data.frame(comparison_result)

gt_table <- gt(comparison_df) %>%
  tab_header(
    title = "Cross-Validation Comparison",
    subtitle = "Poisson model vs. Negative Binomial model"
  )

gt_table

@@

The negative binomial regression model is well-suited for datasets exhibiting overdispersion, which is evidenced in our dataset by various model parameters. This overdispersion signifies that the variance of the dependent variable is larger than would be expected under a Poisson distribution, thus making the negative binomial model a more appropriate choice due to its ability to account for this variability.

@@

Future studies could aim to address these weaknesses by incorporating spatial and demographic variables to account for clustering effects.For instance, additional provinces could be included in the investigation to compare and elucidate factors impacting regional mortality. Future studies can broaden the scope to include more socioeconomical variables, such as diet, lifestyles, environmental pollutants and their interactions. Canada is also home to omany centenarians in the world [@statcan]; conversely, we could compare mortality and the phenomenon of longevity.  Longitudinal studies could offer insights into mortality trends and the long-term effects of environmental factors like climate change.

@@

The raw dataset was downloaded from Statistics Canada open data portal and cleaned by selecting the variables of interest "Year", "Causes of Death", "Characteristics" and "Value". The "Characteristics" variable is converted into a new variable "Death" indicating the number of death attributed to the cause and a new "Rank" variable showing the rank of the cause in a given year. @tbl-6 presents a sample of the cleaned dataset.

```{r}
#| label: tbl-6
#| tbl-cap: "A sample of the cleaned dataset of number of death attributed to causes of death in Canada, 2001-2022"
#| echo: false
#| warning: false
#| message: false
sampled_data <- sample_n(cleaned_data, 5) %>%
  kable()
sampled_data
```
